---
title: "Homework1"
author: "Jonathan Day"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(viridis)
library(babynames)
library(readr)
theme_set(theme_bw())
```

```{r}
columns <- c("status", "ssn", "last_name", "name_suffix", "first_name", "middle_name",
             "vp_code", "death_date", "birth_date", "blanks1", "blanks2",
             "blanks3", "blanks4", "blanks5")

layout <- fwf_widths(c(1, 9, 20, 4, 15, 15, 1, 8, 8, 2, 5, 5, 5, 7),
                     col_names = columns)

ssdm1 <- read_fwf("ssdm1", layout, n_max = 1000000) 
ssdm2 <- read_fwf("ssdm2", layout, n_max = 1000000) 
ssdm3 <- read_fwf("ssdm3", layout, n_max = 1000000) 
```

```{r}
ssdm1 %>% select(first_name, death_date, birth_date) -> ssdm1.cleaned
ssdm2 %>% select(first_name, death_date, birth_date) -> ssdm2.cleaned
ssdm3 %>% select(first_name, death_date, birth_date) -> ssdm3.cleaned
```

```{r}
ssdm <- bind_rows(ssdm1.cleaned,ssdm2.cleaned,ssdm3.cleaned)
```

```{r}
parse_mmddyyyy <- function(date_string) {
  # Replace '00' day with '01'
  date_string <- ifelse(substr(date_string, 3, 4) == "00", 
           paste0(substr(date_string, 1, 2), "01", substr(date_string, 5, 8)),
           date_string)
  mdy(date_string)
}

parse_mmddyyyy(c("12001970", c("01001980")))
```

```{r}
ssdm %>% select(first_name, ends_with("date")) %>% 
    mutate(
        across(ends_with("date"), parse_mmddyyyy),
    ) %>% print ->
    ssdm_cleaned
filter(ssdm_cleaned, year(birth_date) != 0) -> ssdm_cleaned.1
ssdm_cleaned.1$Age <- year(ssdm_cleaned.1$death_date) - year(ssdm_cleaned.1$birth_date)
```

The data I have now is first name, death date, and birth date in a dataframe, in addition to the babynames library I can load gender as well.

For prediction, I need to narrow the focus in on a specific name as input: - obtain list by year of numbers of people with that name alive in that year and how old they are. - From there I can fix the year someone is asking the question (ie its 2010 and I meet a Frank)

-   Then using my list of Franks that are all still alive today I can predict the average age?

To make data wrangling easier I will select a random sample of 100000 for testing ideas

```{r}
ssdm_cleaned.1 %>% filter(first_name == "ROBERT") %>% ggplot(aes(x=year,y=alive)) + geom_density()
```

```{r}
name = 'JONATHAN'

ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2000) %>% 
    drop_na(birth_date, death_date) %>% 
    rowwise() %>% 
    mutate(year = list(year(birth_date):year(death_date))) %>% 
    unnest(year) %>%
    group_by(first_name, year) %>% 
    summarise(alive = n(), .groups = 'drop') %>% 
    ggplot(aes(x=year,y=alive)) + geom_line() + ggtitle(name)

```

```{r}
name = 'ELSIE'

ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2010) %>% 
    drop_na(birth_date, death_date) %>% 
    rowwise() %>% 
    mutate(year = list(year(birth_date):year(death_date))) %>% 
    unnest(year) %>%
    group_by(first_name, year) %>% 
    summarise(alive = n(), .groups = 'drop') %>% 
    ggplot(aes(x=year, y=alive)) + geom_line()
```


```{r}
name = 'ELSIE'

ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2000) %>% 
    drop_na(birth_date, death_date) %>% 
    rowwise() %>% 
    mutate(year = list(year(birth_date):year(death_date))) %>% 
    unnest(year) %>%
    group_by(first_name, year) %>% 
    summarise(alive = n(), .groups = 'drop') %>% 
    ggplot(aes(x=year, y=alive)) + geom_line()
```


```{r}
name = 'MARCELINE'

ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2000) %>% 
    drop_na(birth_date, death_date) %>% 
    rowwise() %>% 
    mutate(year = list(year(birth_date):year(death_date))) %>% 
    unnest(year) %>%
    group_by(first_name, year) %>% 
    summarise(alive = n(), .groups = 'drop') %>% 
    ggplot(aes(x=year, y=alive)) + geom_line()
```
```{r}
library(babynames)

babynames %>% filter(name == "Elsie", sex == "F") %>% ggplot + 
    geom_line(aes(x=year, y=n))
```
```{r}
library(babynames)

babynames %>% filter(name == "Royce", sex == "M") %>% ggplot + 
    geom_line(aes(x=year, y=n))
```
```{r}
set.seed(2020)
top_male_names <- babynames %>%
  filter(sex == "M") %>%
  top_n(500, wt = n) %>%
  sample_n(4)

# Filter the top 500 female names
top_female_names <- babynames %>%
  filter(sex == "F") %>%
  top_n(500, wt = n) %>%
  sample_n(4)

babynames %>% filter(
  (sex == "M" & name %in% top_male_names$name) |
    sex == "F" & name %in% top_female_names$name
) %>% ggplot(aes(x = year, y = prop, group = name, color=name)) + geom_line(size=1) +
  ggtitle("Sample of 4 Male and 4 Female Top 500 Names") +
  xlab("Year") + ylab("Proportion")  +
  scale_color_viridis(discrete = TRUE)
```
```{r}
unique_names <- babynames %>% group_by(year) %>% 
  summarize(unique_names_count = n_distinct(name))
pop_sum <- babynames %>% group_by(year) %>% 
  summarize(population = sum(n))
cor(unique_names$unique_names_count,pop_sum$population)
```

```{r}
set.seed(2025)
name_ex <- babynames %>%
  filter(sex %in% c("M", "F")) %>%
  group_by(sex) %>% top_n(500) %>%
  sample_n(2)

# Filter the babynames dataset for the sampled names
filtered_data <- babynames %>% 
  filter(name %in% name_ex$name)

# Create a ggplot to plot the popularity of the sampled names
ggplot(filtered_data, aes(x = year, y = n, color = interaction(name, sex))) +
  geom_line() +
  labs(title = "Popularity of Baby Names Over Time",
       x = "Year",
       y = "Number of occurrences",
       color = "Name")
```

```{r}
babynames %>%
  summarize(unique_names_count = n_distinct(name))
```



```{r}
read_csv('United States Population by Age and Sex.csv', skip=1) %>% 
    slice(1:101) %>% mutate(
        across(everything(), 
               \(x) str_replace_all(string = x, 
                                    pattern = "[^\\d.]*", 
                                    replacement = "") %>% as.numeric
               )
        ) -> age_structure
```


```{r}
age_structure %>% 
    pivot_longer(cols=2:3, values_to = "Percent of population") %>%
    ggplot(aes(x=Age, y=`Percent of population`, color = name)) +
    geom_line()
```

```{r}
age_structure %>% mutate(birth_year = 2022 - Age) %>% 
    left_join(filter(babynames, name == "John", sex == "M"), 
               by=c("birth_year" = "year")) %>% 
    mutate(p = prop * `Male % of Population`, p = p / sum(p, na.rm = T)) %>% 
    ggplot(aes(x = birth_year, y=p)) + geom_line()
```


```{r}
result <- age_structure %>%
  mutate(birth_year = 2022 - Age) %>%
  left_join(filter(babynames, name == "Jonathan", sex == "M"), 
             by = c("birth_year" = "year")) %>%
  mutate(p = prop * `Male % of Population`, p = p / sum(p, na.rm = TRUE)) %>%
  ggplot(aes(x = birth_year, y = p)) +
  geom_line()

# Extract the index of the maximum value of the vector p
max_index <- which.max(result$data$p)

# Extract the corresponding year
max_year <- result$data$birth_year[max_index]

# Extract the maximum value
max_p <- result$data$p[max_index]

# Print the results
cat("Maximum value of p:", max_p, "\n")
cat("Associated year:", max_year, "\n")

# Compute the variance of the p vector
variance_p <- var(result$data$p, na.rm = TRUE)

# Print the variance
cat("Variance of p:", variance_p, "\n")
```

```{r}
alpha = 0.05; q.alpha = qchisq(1-alpha,1);
q.alpha

```


```{r}
adj_structure <- age_structure %>% mutate(birth_year = 2022 - Age) %>% 
    left_join(filter(babynames, name == "Elsie", sex == "F"), 
               by=c("birth_year" = "year")) %>% 
    mutate(p = prop * `Female % of Population`, p = p / sum(p, na.rm = T)) %>% 
  drop_na()

max.index <- which.max(adj_structure$p)
max.p <- adj_structure$p[max.index]
max.year <- adj_structure$Age[max.index]
chg <- sd(adj_structure$p)
tol <- max.p-chg
output <- adj_structure %>% filter(p > tol) %>% select(Age);
mass <- adj_structure %>% filter(p > tol) %>% summarize(sum(p))
          
cat("Maximum value of p:", max.p, "\n")
cat("Age:", max.year, "\n")
print(mass)
print(output)

ggplot(data=adj_structure) + geom_line(aes(x=Age, y=p)) +
  geom_hline(yintercept=tol, linetype='dashed', color='red') + 
  ggtitle('Age Probability of the Name Elsie') + ylab('Probability')
```
```{r}
adj_structure %>% filter(p > tol) %>% summarise(sum(p))
```


```{r}
kde <- function(x, sampling=seq(-10,10,length.out=100)){
    #get number of points
    n = dim(age_structure)[1]
    #determine bandwidth
    h = (4/3)**0.2 * sd(x) * n**(-0.2)
    #generate the basis for the kde function
    bases = data.frame()
    for (x_i in x) {
      bases = rbind(bases,pnorm(sampling, mean = x_i, sd=h))
    }
    #then sum horizontally
    bases_summed = bases
    #final normalize by the number of points in x
    kde_line = bases_summed/n
    result_list <- list(sampling = sampling, kde_line = kde_line)
    return(result_list)
}
```

```{r}
kde(adj_structure$p,sampling=seq(min(adj_structure$Age),max(adj_structure$Age),length.out=length(adj_structure$Age)))
```


```{r}
library(mclust)
mix_model <- Mclust(age_structure$Age, G = 2)  # Assuming 2 components in the mixture

# Extract the parameters of the fitted mixture model
means <- mix_model$parameters$mean
sds <- sqrt(mix_model$parameters$variance)
weights <- mix_model$parameters$pro
cluster_probs <- matrix(apply(mix_model$z, 1, function(x) sum(x == 1)), ncol = 1)

# Generate a sequence of values for the x-axis
x_seq <- seq(min(age_structure$Age), max(age_structure$Age), length.out = 1000)

# Calculate the mixture model density for each value in the x-axis sequence
mixture_density <- rowSums(outer(x_seq, means, function(x, mu) {
  dnorm(x, mean = mu, sd = sds)
}) * weights * cluster_probs)

# Create a plot of the mixture model KDE
ggplot(data.frame(x = x_seq, density = mixture_density), aes(x = x, y = density)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Mixture Model KDE of Age", x = "Age", y = "Density")
```


```{r}
pred_age <- function(name, G, year_ask) {
  read_csv('United States Population by Age and Sex.csv', skip=1) %>% 
      slice(1:101) %>% mutate(
          across(everything(), 
                 \(x) str_replace_all(string = x, 
                                      pattern = "[^\\d.]*", 
                                      replacement = "") %>% as.numeric
                 )
          ) -> age_structure
  
  adj_structure <- age_structure %>% mutate(birth_year = year_ask - Age) %>% 
      left_join(filter(babynames, name == name, sex == G), 
                 by=c("birth_year" = "year")) %>% 
      mutate(p = prop * `Female % of Population`, p = p / sum(p, na.rm = T)) %>% 
    drop_na()
  
  max.index <- which.max(adj_structure$p)
  max.p <- adj_structure$p[max.index]
  max.year <- adj_structure$Age[max.index]
  chg <- std(adj_structure$p)
  tol <- max.p-chg
  output <- adj_structure %>% filter(p > tol) %>% select(Age)
  
  cat("Maximum value of p:", max.p, "\n")
  cat("Age:", max.year, "\n")
  output
  
  ggplot(data=adj_structure) + geom_line(aes(x=Age, y=p)) + geom_hline(yintercept=tol, color='red')
}
```

```{r}
pred_age('Marceline', 'F', 2010)
```






