---
title: "Homework1"
author: "Jonathan Day"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(viridis)
library(babynames)
library(readr)
library(stringr)
theme_set(theme_bw())
```

```{r}
babynames %>% filter(name == "Elsie", sex == "F") %>% ggplot + 
    geom_line(aes(x=year, y=n))
```

```{r}
library(babynames)

babynames %>% filter(name == "Royce", sex == "M") %>% ggplot + 
    geom_line(aes(x=year, y=n))
```

```{r}
set.seed(2020)
top_male_names <- babynames %>%
  filter(sex == "M") %>%
  top_n(500, wt = n) %>%
  sample_n(4)

# Filter the top 500 female names
top_female_names <- babynames %>%
  filter(sex == "F") %>%
  top_n(500, wt = n) %>%
  sample_n(4)

babynames %>% filter(
  (sex == "M" & name %in% top_male_names$name) |
    sex == "F" & name %in% top_female_names$name
) %>% ggplot(aes(x = year, y = prop, group = name, color=name)) + geom_line(size=1) +
  ggtitle("Sample of 4 Male and 4 Female Top 500 Names") +
  xlab("Year") + ylab("Proportion")  +
  scale_color_viridis(discrete = TRUE)
```

```{r}
unique_names <- babynames %>% group_by(year) %>% 
  summarize(unique_names_count = n_distinct(name))
pop_sum <- babynames %>% group_by(year) %>% 
  summarize(population = sum(n))
cor(unique_names$unique_names_count,pop_sum$population)
```

```{r}
babynames %>%
  summarize(unique_names_count = n_distinct(name))
```



```{r}
read_csv('United States Population by Age and Sex.csv', skip = 1) %>% 
  mutate(across(1:3, ~ as.numeric(str_replace_all(., "[^\\d.]", "")))) %>% 
  slice(1:101) -> age_structure
```


```{r}
age_structure %>% 
    pivot_longer(cols=2:3, values_to = "Percent of population") %>%
    ggplot(aes(x=Age, y=`Percent of population`, color = name)) +
    geom_line()
```

```{r}
age_structure %>% mutate(birth_year = 2022 - Age) %>% 
    left_join(filter(babynames, name == "John", sex == "M"), 
               by=c("birth_year" = "year")) %>% 
    mutate(p = prop * `Male % of Population`, p = p / sum(p, na.rm = T)) %>% 
    ggplot(aes(x = birth_year, y=p)) + geom_line()
```


```{r}
adj_help <- function(fname,gender){
  if (gender == "M"){
    age_structure %>% mutate("birth_year" = 2022 - Age) %>% 
      left_join(filter(babynames, name==fname, sex==gender),by=c("birth_year"="year")) %>%
      mutate(p = prop * `Male % of Population`,
             p = p / sum(p, na.rm = TRUE))-> adj_structure
  } else {
    age_structure %>% mutate("birth_year" = 2022 - Age) %>% 
      left_join(filter(babynames, name==fname,sex==gender),by=c("birth_year"="year")) %>%
      mutate(p = prop * `Female % of Population`,
             p = p / sum(p, na.rm = TRUE))-> adj_structure
  }
  max_index <- which.max(adj_structure$p)
  max_age <- adj_structure$Age[max_index]
  max_p <- adj_structure$p[max_index]
  variance_p <- var(adj_structure$p, na.rm = TRUE)
  chg <- sd(adj_structure$p)
  tol <- max.p-chg
  # output <- adj_structure %>% filter(p > tol) %>% select(Age);
  mass <- adj_structure %>% filter(p > tol) %>% summarize(sum(p))
  return(c(max_age, max_p, variance_p))
}
```

```{r}
adj_help('Mark',"M")
```


``` {r}
# Extract the index of the maximum value of the vector p
max_index <- which.max(result$data$p)

# Extract the corresponding year
max_year <- result$data$birth_year[max_index]

# Extract the maximum value
max_p <- result$data$p[max_index]
# Compute the variance of the p vector
variance_p <- var(result$data$p, na.rm = TRUE)
# Print the results
cat("Maximum value of p:", max_p, "\n")
cat("Associated year:", max_year, "\n")
# Print the variance
cat("Variance of p:", variance_p, "\n")
```

```{r}
alpha = 0.05; q.alpha = qchisq(1-alpha,1);
q.alpha

```


```{r}
adj_help <- function(name, gender){
  gen_col <- if(gender=='M'){
      age_structure$`Male % of Population`
  } else{ age_structure$`Female % of Population`
    }
  adj_structure <- age_structure %>% 
    mutate(birth_year = 2022 - Age) %>% 
    left_join(
      filter(babynames, name == "name", sex == gender), 
      by = c("birth_year" = "year")
    ) %>% 
  mutate(p = prop * gen_col,
         p = p / sum(p, na.rm = TRUE))
  return(adj_structure)
}
```

```{r}
print(adj_help('Marcus', 'M'))
```


```{r}
max.index <- which.max(adj_structure$p)
max.p <- adj_structure$p[max.index]
max.year <- adj_structure$Age[max.index]
chg <- sd(adj_structure$p)
tol <- max.p-chg
output <- adj_structure %>% filter(p > tol) %>% select(Age);
mass <- adj_structure %>% filter(p > tol) %>% summarize(sum(p))
          
cat("Maximum value of p:", max.p, "\n")
cat("Age:", max.year, "\n")
print(mass)
print(output)

ggplot(data=adj_structure) + geom_line(aes(x=Age, y=p)) +
  geom_hline(yintercept=tol, linetype='dashed', color='red') + 
  ggtitle('Age Probability of the Name Elsie') + ylab('Probability')
```
```{r}
max.index <- which.max(adj_structure$p)
max.p <- adj_structure$p[max.index]
max.year <- adj_structure$Age[max.index]
chg <- sd(adj_structure$p)
tol <- max.p-chg
output <- adj_structure %>% filter(p > tol) %>% select(Age);
mass <- adj_structure %>% filter(p > tol) %>% summarize(sum(p))
          
cat("Maximum value of p:", max.p, "\n")
cat("Age:", max.year, "\n")
print(mass)
print(output)

ggplot(data=adj_structure) + geom_line(aes(x=Age, y=p)) +
  geom_hline(yintercept=tol, linetype='dashed', color='red') + 
  ggtitle('Age Probability of the Name Elsie') + ylab('Probability')
```


```{r}
adj_structure %>% filter(p > tol) %>% summarise(sum(p))
```

KDE Attempts
```{r}
kde <- function(x, sampling=seq(-10,10,length.out=100)){
    #get number of points
    n = dim(age_structure)[1]
    #determine bandwidth
    h = (4/3)**0.2 * sd(x) * n**(-0.2)
    #generate the basis for the kde function
    bases = data.frame()
    for (x_i in x) {
      bases = rbind(bases,pnorm(sampling, mean = x_i, sd=h))
    }
    #then sum horizontally
    bases_summed = bases
    #final normalize by the number of points in x
    kde_line = bases_summed/n
    result_list <- list(sampling = sampling, kde_line = kde_line)
    return(result_list)
}
```

```{r}
kde(adj_structure$p,sampling=seq(min(adj_structure$Age),max(adj_structure$Age),length.out=length(adj_structure$Age)))
```

```{r}
library(mclust)
mix_model <- Mclust(age_structure$Age, G = 2)  # Assuming 2 components in the mixture

# Extract the parameters of the fitted mixture model
means <- mix_model$parameters$mean
sds <- sqrt(mix_model$parameters$variance)
weights <- mix_model$parameters$pro
cluster_probs <- matrix(apply(mix_model$z, 1, function(x) sum(x == 1)), ncol = 1)

# Generate a sequence of values for the x-axis
x_seq <- seq(min(age_structure$Age), max(age_structure$Age), length.out = 1000)

# Calculate the mixture model density for each value in the x-axis sequence
mixture_density <- rowSums(outer(x_seq, means, function(x, mu) {
  dnorm(x, mean = mu, sd = sds)
}) * weights * cluster_probs)

# Create a plot of the mixture model KDE
ggplot(data.frame(x = x_seq, density = mixture_density), aes(x = x, y = density)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Mixture Model KDE of Age", x = "Age", y = "Density")
```

Working on function to automate:
```{r}
pred_age <- function(name, G, year_ask) {
  read_csv('United States Population by Age and Sex.csv', skip=1) %>% 
      slice(1:101) %>% mutate(
          across(everything(), 
                 \(x) str_replace_all(string = x, 
                                      pattern = "[^\\d.]*", 
                                      replacement = "") %>% as.numeric
                 )
          ) -> age_structure
  
  adj_structure <- age_structure %>% mutate(birth_year = year_ask - Age) %>% 
      left_join(filter(babynames, name == name, sex == G), 
                 by=c("birth_year" = "year")) %>% 
      mutate(p = prop * `Female % of Population`, p = p / sum(p, na.rm = T)) %>% 
    drop_na()
  
  max.index <- which.max(adj_structure$p)
  max.p <- adj_structure$p[max.index]
  max.year <- adj_structure$Age[max.index]
  chg <- std(adj_structure$p)
  tol <- max.p-chg
  output <- adj_structure %>% filter(p > tol) %>% select(Age)
  
  cat("Maximum value of p:", max.p, "\n")
  cat("Age:", max.year, "\n")
  output
  
  print(ggplot(data=adj_structure) + geom_line(aes(x=Age, y=p)) +
          geom_hline(yintercept=tol, color='red'))
}
```

```{r}
p_age <- function(name, G){
  read_csv('United States Population by Age and Sex.csv', skip=1) %>% 
      slice(1:101) %>% mutate(
          across(everything(), 
                 \(x) str_replace_all(string = x, 
                                      pattern = "[^\\d.]*", 
                                      replacement = "") %>% as.numeric
                 )
          ) -> age_structure
  
  adj_structure <- age_structure %>% mutate(birth_year = 2022 - Age) %>% 
      left_join(filter(babynames, name == name, sex == G), 
                 by=c("birth_year" = "year")) %>% 
      mutate(p = prop * `Female % of Population`, p = p / sum(p, na.rm = T)) %>% 
    drop_na()

  ggplot(data=adj_structure) + geom_line(aes(x=Age, y=p)) +
    geom_hline(yintercept=tol, linetype='dashed', color='red') + 
    ggtitle('Age Probability of the Name Elsie') + ylab('Probability')
}
```

```{r}
p_age('Marceline', 'F')
```
```
```{r}
read_csv('United States Population by Age and Sex.csv', skip=1) %>% 
    slice(1:101) %>% mutate(
        across(everything(), 
               \(x) str_replace_all(string = x, 
                                    pattern = "[^\\d.]*", 
                                    replacement = "") %>% as.numeric
               )
        ) -> age_structure
```











Death Master File Stuff
```{r}
columns <- c("status", "ssn", "last_name", "name_suffix", "first_name", "middle_name",
             "vp_code", "death_date", "birth_date", "blanks1", "blanks2",
             "blanks3", "blanks4", "blanks5")

layout <- fwf_widths(c(1, 9, 20, 4, 15, 15, 1, 8, 8, 2, 5, 5, 5, 7),
                     col_names = columns)

# ssdm1 <- read_fwf("ssdm1", layout, n_max = 1000000) 
# ssdm2 <- read_fwf("ssdm2", layout, n_max = 1000000) 
# ssdm3 <- read_fwf("ssdm3", layout, n_max = 1000000) 
```

```{r}
#ssdm1 %>% select(first_name, death_date, birth_date) -> ssdm1.cleaned
#ssdm2 %>% select(first_name, death_date, birth_date) -> ssdm2.cleaned
#ssdm3 %>% select(first_name, death_date, birth_date) -> ssdm3.cleaned
```

```{r}
#ssdm <- bind_rows(ssdm1.cleaned,ssdm2.cleaned,ssdm3.cleaned)
```

```{r}
parse_mmddyyyy <- function(date_string) {
  # Replace '00' day with '01'
  #date_string <- ifelse(substr(date_string, 3, 4) == "00", 
  #         paste0(substr(date_string, 1, 2), "01", substr(date_string, 5, 8)),
  #         date_string)
  #mdy(date_string)
}

parse_mmddyyyy(c("12001970", c("01001980")))
```

```{r}
#ssdm %>% select(first_name, ends_with("date")) %>% 
    #mutate(
    ##    across(ends_with("date"), parse_mmddyyyy),
    #) %>% print ->
    #ssdm_cleaned
#filter(ssdm_cleaned, year(birth_date) != 0) -> ssdm_cleaned.1
#ssdm_cleaned.1$Age <- year(ssdm_cleaned.1$death_date) - year(ssdm_cleaned.1$birth_date)
```

```{r}
# ssdm_cleaned.1 %>% filter(first_name == "ROBERT") %>% ggplot(aes(x=year,y=alive)) + geom_density()
```

```{r}
#name = 'JONATHAN'

# ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2000) %>% 
#    drop_na(birth_date, death_date) %>% 
#    rowwise() %>% 
#    mutate(year = list(year(birth_date):year(death_date))) %>% 
#    unnest(year) %>%
#    group_by(first_name, year) %>% 
#    summarise(alive = n(), .groups = 'drop') %>% 
#    ggplot(aes(x=year,y=alive)) + geom_line() + ggtitle(name)

```

```{r}
#name = 'ELSIE'

#ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2010) %>% 
#    drop_na(birth_date, death_date) %>% 
#    rowwise() %>% 
#    mutate(year = list(year(birth_date):year(death_date))) %>% 
#    unnest(year) %>%
#    group_by(first_name, year) %>% 
#    summarise(alive = n(), .groups = 'drop') %>% 
#    ggplot(aes(x=year, y=alive)) + geom_line()
```


```{r}
#name = 'ELSIE'

#ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2000) %>% 
#    drop_na(birth_date, death_date) %>% 
#    rowwise() %>% 
#    mutate(year = list(year(birth_date):year(death_date))) %>% 
#    unnest(year) %>%
#    group_by(first_name, year) %>% 
#    summarise(alive = n(), .groups = 'drop') %>% 
#    ggplot(aes(x=year, y=alive)) + geom_line()
```


```{r}
#name = 'MARCELINE'

#ssdm_cleaned.1 %>% filter(first_name == name, year(death_date)>2000) %>% 
#    drop_na(birth_date, death_date) %>% 
#    rowwise() %>% 
#    mutate(year = list(year(birth_date):year(death_date))) %>% 
#    unnest(year) %>%
#    group_by(first_name, year) %>% 
#    summarise(alive = n(), .groups = 'drop') %>% 
#    ggplot(aes(x=year, y=alive)) + geom_line()
```




